---
title: "COMPAS Data Visualization"
author: "Lucius Bynum, Preethi Seshadri"
date: '19 November 2017'
# output: 
#   html_document:
#     theme: journal
#     highlight: tango
#     df_print: paged
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE)
options(digits=4)

library(dplyr)
library(GGally)
library(ggplot2)
library(e1071)
```

# Introduction

# Problem Statement

# Data Description

ProPublica collected data on **TODO**

In the data set we look at, they considered only people who either "recidivated within two years of their crime or "recidivated in tow years, or had at least tow years outside of a correctional facility."

### Data Dictionary

ProPublica did not provide a data dictionary explaining their variables. Through some manual exploration, we came up with the following descriptions for our best guess at what each variable measures.

<!-- -id: unque identifier for each individual -->
<!-- -name: first and last name -->
<!-- -first: first name -->
<!-- -last: last name -->
<!-- -compas_screening_dat: date on which decile_score was given -->
<!-- -sex: sex (male or female) -->
<!-- -dob: date of birth -->
<!-- -age: age in years -->
<!-- -age_cat: age category (less than 25, 25-45, greater than 45) -->
<!-- -race: race (African-American, Asian, Caucasian, Hispanic, Native American, Other) -->
<!-- -juv_fel_count: juvenile felony count -->
<!-- -decile_score: COMPAS Risk of Recidivism score from 1 to 10 -->
<!-- -juv_misd_count: juvenile misdemeanor count -->
<!-- -juv_other_count: juvenile other offenses count -->
<!-- -priors_count: prior offenses count -->
<!-- -days_b_screening_arrest: number of days between COMPAS screening and arrest **TODO**: negative values? -->
<!-- -c_jail_in: jail entry date for original crime -->
<!-- -c_jail_out: jail exit date for original crime -->
<!-- -c_case_number: case number for original crime -->
<!-- -c_offense_date: offense date of original crime -->
<!-- -c_arrest_date: arrest date for original crime -->
<!-- -c_days_from_compas: days between COMPAS screening and original crime offense date -->
<!-- -c_charge_degree: charge degree of original crime -->
<!-- -c_charge_desc: description of charge for original crime -->
<!-- -is_recid: binary indicator of recidivation (1=individual recidivated, 0=individual did not recidivate) -->
<!-- -r_case_number: case number of follow-up crime -->
<!-- -r_charge_degree: charge degree of follow-up crime -->
<!-- -r_days_from_arrest: number of days between follow-up crime and arrest date **TODO**: why negative value here? -->
<!-- -r_offense_date: date of follow-up crime -->
<!-- -r_charge_desc: description of charge for follow-up crime -->
<!-- -r_jail_in: jail entry date for follow-up crime -->
<!-- -r_jail_out: jail exit date for follow-up crime -->
<!-- -violent_recid: values are all `NA`. This column is ignored. -->
<!-- -is_voilent_recid: binary indicator of violent follow-up crime (1=follow-up crime was violent, 0=follow-up crime was non-violent) -->
<!-- -vr_case_number: case number for violent follow-up crime -->
<!-- -vr_charge_degree: charge degree for violent follow-up crime -->
<!-- -vr_offense_date: date of offense for violent follow-up crime -->
<!-- -vr_charge_desc: description of charge for violent follow-up crime -->
<!-- -type_of_assessment: the type of COMPAS score given for decile_score (here all values are Risk of Recidivism) -->
<!-- -decile_score.1: repeat column of decile_score -->
<!-- -score_text: ProPublica-defined category of decile_score (High: 8-10, Medium: 5-7, Low: 1-4) -->
<!-- -screening_date: repeat column of compas_screening_date  -->
<!-- -v_type_of_assessment: the type of COMPAS score given for v_decile_score (here all values are Risk_of_Violence) -->
<!-- -v_decile_score: COMPAS Risk of Violence score from 1 to 10 -->
<!-- -v_score_text: ProPublica-defined category of v_decile_score (High: 8-10, Medium: 5-7, Low: 1-4) -->
<!-- -v_screening_date: date on which v_decile_score was given -->
<!-- -in_custody: date on which individual was brought into custody -->
<!-- -out_custody: date on which individual was released from custody -->
<!-- -priors_count.1: repeat column of priors_count -->
<!-- -start: **TODO** -->
<!-- -end: **TODO** -->
<!-- -event: **TODO** -->

| Variable| Description|
|--------------------------|----------------------------------------------------------------------------------------------------------------|
| id                      | unque identifier for each individual|
| name                    | first and last name|
| first                   | first name|
| last                    | last name|
| compas_screening_dat    | date on which decile_score was given|
| sex                     | sex (male or female)|
| dob                     | date of birth|
| age                     | age in years|
| age_cat                 | age category (less than 25, 25-45, greater than 45)|
| race                    | race (African-American, Asian, Caucasian, Hispanic, Native American, Other)|
| juv_fel_count           | juvenile felony count|
| decile_score            | COMPAS Risk of Recidivism score from 1 to 10|
| juv_misd_count          | juvenile misdemeanor count|
| juv_other_count         | juvenile other offenses count|
| priors_count            | prior offenses count|
| days_b_screening_arrest | number of days between COMPAS screening and arrest **TODO** negative values?|
| c_jail_in               | jail entry date for original crime|
| c_jail_out              | jail exit date for original crime|
| c_case_number           | case number for original crime|
| c_offense_date          | offense date of original crime|
| c_arrest_date           | arrest date for original crime|
| c_days_from_compas      | days between COMPAS screening and original crime offense date|
| c_charge_degree         | charge degree of original crime|
| c_charge_desc           | description of charge for original crime|
| is_recid                | binary indicator of recidivation (1=individual recidivated, 0=individual did not recidivate)|
| r_case_number           | case number of follow-up crime|
| r_charge_degree         | charge degree of follow-up crime|
| r_days_from_arrest      | number of days between follow-up crime and arrest date **TODO** why negative value here?|
| r_offense_date          | date of follow-up crime|
| r_charge_desc           | description of charge for follow-up crime|
| r_jail_in               | jail entry date for follow-up crime|
| r_jail_out              | jail exit date for follow-up crime|
| violent_recid           | values are all `NA`. This column is ignored.|
| is_voilent_recid        | binary indicator of violent follow-up crime (1=follow-up crime was violent, 0=follow-up crime was non-violent) |
| vr_case_number          | case number for violent follow-up crime|
| vr_charge_degree        | charge degree for violent follow-up crime|
| vr_offense_date         | date of offense for violent follow-up crime|
| vr_charge_desc          | description of charge for violent follow-up crime|
| type_of_assessment      | the type of COMPAS score given for decile_score (here all values are Risk of Recidivism)|
| decile_score.1          | repeat column of decile_score|
| score_text              | ProPublica-defined category of decile_score (High=8-10, Medium=5-7, Low=1-4)|
| screening_date          | repeat column of compas_screening_date|
| v_type_of_assessment    | the type of COMPAS score given for v_decile_score (here all values are Risk_of_Violence)|
| v_decile_score          | COMPAS Risk of Violence score from 1 to 10|
| v_score_text            | ProPublica-defined category of v_decile_score (High=8-10, Medium=5-7, Low=1-4)|
| v_screening_date        | date on which v_decile_score was given|
| in_custody              | date on which individual was brought into custody|
| out_custody             | date on which individual was released from custody|
| priors_count.1          | repeat column of priors_count|
| start                   | **TODO**|
| end                     | **TODO**|
| event                   | **TODO**|

ProPublica obtained this data with the goal of analyzing Northpointe Inc.’s commercial recidivism modeling tool -- COMPAS. Aggregating data from public records, they collected data on 18,610 individuals who received COMPAS scores from 2013 to 2014, including demographic information, public criminal records, and incarceration records.

*How are COMPAS scores used?*

ProPublica describes that at least in Broward County, they “primarily [use] the score to determine whether to release or detain a defendant before his or her trial.” 11,757 of the individuals in the database had their COMPAS scores used to assess whether or not they should be released before their trial.

*What are COMPAS scores?*

There are three types of COMPAS score. Each measures a type of ‘risk’ associated with a criminal re-offending in some way on a scale of 1 (low) to 10 (high). As ProPublica describes, these include 

- *Risk of Recidivism*: ProPublica defines this as the person in question committing a “criminal offense that [results] in a jail booking and [takes] place after the crime for which the person was COMPAS scored.”  Northpointe hopes to use this score to predict “a new misdemeanor or felony offense within two years of the COMPAS administration date.”
- *Risk of Violence*: They use the FBI definition of violent crime: 
    
    *In the FBI’s Uniform Crime Reporting (UCR) Program, violent crime is composed of four offenses: murder and nonnegligent manslaughter, forcible rape, robbery, and aggravated assault. Violent crimes are defined in the UCR Program as those offenses which involve force or threat of force. - ucr.fbi.gov*

- *Risk of Failure to Appear*: As evidenced by the name, this describes a failure to appear at the court hearing. 



# Exploratory Data Analysis

We first load in the data provided by ProPublica, and take a quick look at summaries of each variable to get a sense of distribution and potential outliers. We'll look at half of the columns at a time to keep things manageable. 

```{r}
compas_scores <- read.csv('compas-scores-two-years.csv')
summary(compas_scores[1:(ncol(compas_scores)/2)])
```

Here we notice there may be large outliers in many of the crime count variables, such as `juv_fel_count`, `juv_misd_count`, `juv_other_count`, and `priors_count`. We expect these are simply accurate observations corresponding to individuals with high numbers of prior offenses. Thus we will not remove these individuals from the data but will be aware of them as potential influential points when we later fit any models. We also note that the values for the 'days_from' variables are quite variable which may be relevant if we use those variables in later analysis. Looking at the second half of the columns, we have:

```{r}
summary(compas_scores[(ncol(compas_scores)/2 + 1):ncol(compas_scores)])
```

With the second half we have similar characteristics as before. We remove the `violent_recid` column given that all values are `NA` (as mentioned in the data dictionary). Apart from that column, we make no other changes.

```{r}
vars <- compas_scores %>% 
  select(sex, race, age, ends_with('count'), c_charge_degree, is_recid)

# ggplot(vars, aes(x = race, y = decile_score)) + geom_boxplot()
```

## ProPublica's Bias-Assessment Model

This section is meant to recreate ProPublica's logistic regression model for assessing bias in COMPAS scores. The code and explanations in this section are taken from their [published methodlogy and code](https://github.com/propublica/compas-analysis). 

Using the same data set, they first filter rows based on the following criteria.

1. consider only individuals with a COMPAS score
2. assure the COMPAS score corresponds to the correct crime i.e. the score was given within 30 days of the arrest
3. do not include ordinary traffic offenses

Next we use their code to perform this filtering:

```{r}
# code from https://github.com/propublica/compas-analysis
df <- compas_scores %>%
  select(age, c_charge_degree, race, age_cat, score_text, sex, priors_count, days_b_screening_arrest, decile_score, is_recid, two_year_recid, c_jail_in, c_jail_out) %>% 
  filter(days_b_screening_arrest <= 30) %>%
  filter(days_b_screening_arrest >= -30) %>%
  filter(is_recid != -1) %>%
  filter(c_charge_degree != "O") %>%
  filter(score_text != 'N/A')
nrow(df)
```

In order to use this data to assess racial bias in scoring, the ProPublica analysts first create several factor variables from the existing columns.

```{r}
# code from https://github.com/propublica/compas-analysis
df <- mutate(df, crime_factor = factor(c_charge_degree)) %>%
      mutate(age_factor = as.factor(age_cat)) %>%
      within(age_factor <- relevel(age_factor, ref = 1)) %>%
      mutate(race_factor = factor(race)) %>%
      within(race_factor <- relevel(race_factor, ref = 3)) %>%
      mutate(gender_factor = factor(sex, labels= c("Female","Male"))) %>%
      within(gender_factor <- relevel(gender_factor, ref = 2)) %>%
      mutate(score_factor = factor(score_text != "Low", labels = c("LowScore","HighScore")))
```

For the `age_factor` they make "25 - 45" the reference level, for `race_factor` "Caucasian" is the reference level, for `gender_factor` "Male" is the reference level. 

Next they fit a logistic regression model to predict `score_factor` from the other variables.

```{r}
# code from https://github.com/propublica/compas-analysis
pp_model <- glm(
  score_factor ~ gender_factor + age_factor + race_factor + priors_count + crime_factor + two_year_recid, 
  family="binomial", 
  data=df
)
summary(pp_model)
```

#### Calculating Relative Risk by Demographic

Next they compute how much more likely different demographic groups are to receive a higher score than others. The logistic regression model allows us to measure this difference after correcting for the other variables included in the model. The quantity ProPublica uses to compare black defendants to white defendants (or men to women, old defendants to young defendants, etc.) is called **relative risk**. ProPublica does not explain where this quantity comes from in their analysis, so we'll provide some quick background on logistic regression to justify the calculation. The following explanation is inspired by USC professor Sandy Eckel's slides [here](http://www-hsc.usc.edu/~eckel/biostat2/notes/notes14.pdf).

Logistic regression models a linear relationship between the log odds ratio for the probability of interest and the given predictor variables. An odds ratio measures the odds of success

$$\texttt{odds ratio} = \frac{\texttt{probability of success}}{\texttt{probability of failure}} = \frac{P}{1 - P}$$

where $P$ is the probability of success. The log odds ratio is simply the log of this quantity. Thus the logistic regression model for observation $x_i$ is

$$\log(\frac{P_{x_i}}{1-P_{x_i}}) = \beta_0 + \beta_1x_{i1} + \ldots + \beta_px_{ip}$$

where the probability of success for $x_i$ is $P_{x_i}$, and we have $p$ predictors with corresponding coeffecients $\beta_j$ and observed values $x_{ij}$ for $j=1,\ldots,p$. Given that our model here uses categorical predictors (factors), the coeffecients we estimate give us the **change in log odds** for the corresponding variable. Thus if we let $P_{x_i}$ be the probability that individual $x_i$ gets a high COMPAS score, then with coefficent $\beta_1$ for `gender_factor`, we would have

$$\beta_0: \text{the log odds of getting a high COMPAS score for men}$$
$$\beta_1: \text{the difference in log odds of getting a high COMPAS score for women compared to men}$$

The important observation here is that because men are the reference level for the `gender_factor` categorical variable, $\beta_1$ measures a **difference** relative to men. Thus if we want to answer the question, "How much more likely are women to get a high COMPAS score than men?" we'll want to use

$$\beta_0 + \beta_1: \text{the log odds of getting a high COMPAS score for women}$$

to get the comparison. One other observation will also be helpful to calculate relative risk. We solve for $P_{x_i}$ as follows.

$$\log(\frac{P}{1 - P}) = x \rightarrow P = \frac{e^x}{1 + e^x} = \texttt{sigmoid}(x)$$

Thus we calculate relative risk for the categorical variable corresponding to $\beta_1$ as:

$$\texttt{relative risk} = \frac{P_1}{P_2} = \frac{\texttt{sigmoid}(\beta_0 + \beta_1)}{\texttt{sigmoid}(\beta_0)}$$

ProPublica computes relative risk to compare blacks to whites, men to women, and people under 25 to middle-aged people in terms of COMPAS scores. They get the following results.

```{r}
# code adapted from https://github.com/propublica/compas-analysis
model_intercept <- coef(pp_model)['(Intercept)']
black_coef <- coef(pp_model)['race_factorAfrican-American']
(relative_risk_black_v_white <- sigmoid(model_intercept + black_coef) / sigmoid(model_intercept))
```

As ProPublica states, this shows us that "Black defendants are 45% more likely than white defendants to receive a higher [COMPAS] score correcting for the seriousness of their crime, previous arrests, and future criminal behavior." Similarly, women are 19.4\% more likely than men and people under 25 are 2.5 times as likely as middle aged people to get a higher score:

```{r}
# code adapted from https://github.com/propublica/compas-analysis
woman_coef <- coef(pp_model)['gender_factorFemale']
(relative_risk_woman_v_man <- sigmoid(model_intercept + woman_coef) / sigmoid(model_intercept))
```

```{r}
# code adapted from https://github.com/propublica/compas-analysis
age_coef <- coef(pp_model)['age_factorLess than 25']
(relative_risk_young_v_middleage <- sigmoid(model_intercept + age_coef) / sigmoid(model_intercept))
```

## Our Logistic Regression Model

```{r}
regmod <- glm(is_recid ~., family = binomial(link='logit'), data = vars)
summary(regmod)
fitted_values <- fitted(regmod)
vars <- cbind(vars, fitted_values)
```

## Using ProPublica's Bias-Assessment Model on Our Model

We can see if our model suffers from the same bias using ProPublica's bias-assessment methodology. We'll set up the same model they did, but this time instead of predicting `score_factor` for the COMPAS score, we'll create a `score_factor` variable *from our model's predictions*.

## Variable Selection

# Clustering





